Beyond the Arc: The Impact of  3 point Shooting in the Modern NBA

### 2024-04-18 paper/paper.qmd
Final changes

---
title: "Beyond the Arc: How 3 point shooting has impacted winning in the modern NBA"
subtitle: "Analysing the relationship between a team's 3 point shooting and sucess in the regular season"
author: 
  - Aryaman Suri
thanks: "Code and data are available at: https://github.com/aryamansuri/NBA-3PointAnalysis"
date: today
date-format: long
abstract: "In recent years, the three-point shot has emerged as a pivotal aspect of basketball strategy in the NBA. This study uses statistical modelling tools to analyze the association between three-point shooting and winning in the current NBA. Using season-wise NBA statistics, I examine the cumulative effect of teams' three-point shooting on their overall victory total. This study quantifies this link using Generalized Linear Models, which take into account parameters such as the amount of three-point shots made and other pertinent variables. The findings exemplify the importance of three-point shooting as an NBA success factor. I examine the ramifications of the findings for team strategy, player development, and basketball analytics. Additionally, ethical concerns and study limitations are addressed."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The three point line was introduced in the NBA in the 1979-80 season and has brought unprecedented balance to the game. Instead of the game being centered around the big man or athletic wings, the 3 point shot gave everybody on the team a new powerful weapon on offenses. Today, there are 230 players who averaged at least one made three pointer in the 2022-23 regular season. The estimand of this paper is to see if these 3 point shots help the team win games in the regular season.

This paper aims to estimate the association between three-point shooting measures and win probability using advanced statistical techniques after meticulously analyzing season-wise data. The findings highlighted the importance of three-point shooting efficiency as a predictor of team performance in modern basketball and how important it is to be a good three point shooting team for success in today's game.

Understanding this link has far-reaching ramifications for numerous parties, including coaches, players, analysts, and team administration. This study adds to the ongoing debate in the NBA about strategic decision-making and player development by explaining the significance of three-point shooting in deciding game outcomes. Essentially, the question "Does three point shooting actually lead to team success?", is being asked.

I begin by exploring the underlying data, which will provide insights into the variables and metrics used in our analyses. Following that, I explain the complexities of the modelling approach, describing the reasoning behind the statistical techniques and justifying the methodology. Later, the results of the analysis are discussed, diving into the empirical findings and their implications for understanding the relationship between three-point shooting efficiency and NBA team success. Then  I have included a thorough conversation, contextualizing the findings within the larger landscape of basketball analytics and emphasizing significant takeaways from this research. Finally, I summarize the findings and propose paths for future research to further the understanding of this dynamic interaction.

# Data

In this study, I utilized data sourced from the NBA's official database (@cite-NBA) to examine the relationship between various team performance metrics and their corresponding winning percentages throughout the regular season. My analysis aims to uncover insights into the factors contributing to team success in the NBA, specifically 3 point shooting. The NBA's database serves as a complete repository of statistical information on teams, players, and games, providing valuable insights into the dynamics of professional basketball. Below, I provide a detailed overview of the data collection process, cleaning procedures, and subsequent analysis.

@@

While the measurement of this model's data can be criticized by NBA fans , we've deliberately opted for **3PM** over **3PA** and **3P%** for several reasons. **3PM** serves as a robust metric, capturing a team's volume of successful three-point shots with reasonable efficiency. Unlike **3PA**, which might bias the model towards teams overly reliant on three-point attempts irrespective of their success rate, **3PM** provides a balanced representation of both volume and efficiency in three-point shooting. Similarly, **3P%** could skew the analysis towards teams with lower volumes of three-point attempts, overlooking the broader spectrum of team performance.

By prioritizing **3PM** alongside Wins, our model aims to uncover nuanced insights into the impact of three-point scoring on overall team success while mitigating potential biases associated with alternative metrics. This strategic selection ensures a throrough exploration of the relationship between three-point performance and winning outcomes in NBA games.

# Model

@@

I hypothesize a positive relationship between the number of 3-point field goals made by a team and their number of wins throughout the regular NBA season. This hypothesis is rooted in the understanding that teams with higher 3-point shooting efficiency typically demonstrate stronger offensive capabilities, which often translate into higher-scoring games and, consequently, more victories. This relationship is particularly pertinent in today's NBA landscape, where teams increasingly prioritize offensive strategies and exploit the scoring potential offered by 3-point shots.

The chosen Bayesian analysis model (@cite-rstanarm) aims to quantitatively assess the impact of 3-point field goals made on team wins and to provide insights into the significance of this performance metric in determining overall team success. By modeling the relationship between 3-point field goals made and wins using a normal distribution for the response variable, we acknowledge the continuous nature of team wins and seek to capture the variability in win counts across different levels of 3-point shooting proficiency. This approach allows for a nuanced understanding of how variations in 3-point performance contribute to variations in team outcomes, thereby elucidating the importance of 3-point shooting in the context of NBA team success.

# Results

@@

```{r}
#| echo: false
#| label: fig-2
#| fig-cap: "Graph depicting the **Wins** vs the **3 Points Made** per game"
#| warning: false
ggplot(cleaned_data, aes(x = X3PM, y = W)) +
  geom_point(color = "orange") +
  labs(x = "3PM", y = "Wins", title = "Wins vs 3PM") +
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.grid.major = element_blank(),         # Remove major gridlines
    panel.grid.minor = element_blank()          # Remove minor gridlines
  )
```

@@

team_colors <- c(
  "Atlanta Hawks" = "#E03A3E",
  "Boston Celtics" = "#007A33",
  "Brooklyn Nets" = "#000000",
  "Charlotte Hornets" = "#00788C",
  "Chicago Bulls" = "#CE1141",
  "Cleveland Cavaliers" = "#860038",
  "Dallas Mavericks" = "#00538C",
  "Denver Nuggets" = "#0E2240",
  "Detroit Pistons" = "#C8102E",
  "Golden State Warriors" = "#1D428A",
  "Houston Rockets" = "#CE1141",
  "Indiana Pacers" = "#002D62",
  "LA Clippers" = "#C8102E",
  "Los Angeles Lakers" = "#552583",
  "Memphis Grizzlies" = "#5D76A9",
  "Miami Heat" = "#98002E",
  "Milwaukee Bucks" = "#00471B",
  "Minnesota Timberwolves" = "#0C2340",
  "New Orleans Pelicans" = "#0C2340",
  "New York Knicks" = "#006BB6",
  "Oklahoma City Thunder" = "#007AC1",
  "Orlando Magic" = "#0077C0",
  "Philadelphia 76ers" = "#006BB6",
  "Phoenix Suns" = "#1D1160",
  "Portland Trail Blazers" = "#E03A3E",
  "Sacramento Kings" = "#5A2D81",
  "San Antonio Spurs" = "#C4CED4",
  "Toronto Raptors" = "#CE1141",
  "Utah Jazz" = "#002B5C",
  "Washington Wizards" = "#002B5C"
)
ggplot(grouped_data, aes(x = Average_3PM, y = reorder(TEAM, Average_3PM), fill = TEAM)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Average_3PM, 1)), hjust = 1.2, size = 3, color = "black") +
  labs(x = "Average 3-Pointers Made", y = "Team") +
  scale_fill_manual(values = team_colors) +
  theme(legend.position = "none")
```
@fig-3 shows the average 3 point shots made per game across the 8 seasons that are involved in our study.
```{r}
#| echo: false
#| label: fig-1
#| fig-cap: "Graph depicting the **Average team wins** and **predicted team wins** vs **Average three point shots made** per game"
#| warning: false
# Calculate predicted wins
predicted_wins <- fitted(nba_model2)

@@

# Discussion

In this study, we aimed to investigate the relationship between three-point shooting efficiency and team success in the NBA. Our analysis involved an thorough examination of season-wise NBA statistics, focusing on variables such as three-pointers made (3PM) and team wins (W). We employed advanced statistical techniques, including data cleaning, correlation analysis, and regression modeling, to uncover insights into the impact of three-point shooting on team performance. By analyzing the data and modeling the relationship between 3PM and W, we sought to show the significance of three-point shooting as a predictor of NBA team success.

Our analysis revealed a compelling association between three-point shooting proficiency and team success in the NBA. We found that teams with higher rates of three-pointers made per game tended to achieve more wins over the course of the regular season. This suggests that three-point shooting plays a key role in determining the outcome of NBA games, with proficient long-range shooters contributing significantly to their team's overall success. These findings underscore the strategic importance of three-point shooting in modern basketball and highlight its impact on team performance.
Additionally, our analysis uncovered nuanced insights into the dynamics of three-point shooting and team success. We observed variations in the relationship between 3PM and W across different teams and seasons, indicating that the impact of three-point shooting may vary based on contextual factors such as team composition, playing style, and opponent strategies. Furthermore, our findings suggest that while three-point shooting is a significant predictor of team success, it is not the sole determinant, with other factors such as defense, rebounding, and overall offensive efficiency also playing pivotal roles in determining game outcomes.

@@

Furthermore, while our modeling approach accounted for various factors, including team performance metrics and game outcomes, it may not capture all relevant variables that could impact the relationship between three-point shooting and team success. Success in the NBA cannot simply be boiled down to efficient volume three point shooting since it is a nuanced game with many other factors contributing to a team's overall success.

@@

In essence, while the three-point shot remains a potent weapon in the modern NBA, its effectiveness is contingent upon a nuanced understanding of its role within a team's overall offensive strategy. As the Houston Rockets learned, basketball is not solely a game of statistics—factors such as matchup dynamics, defensive schemes, and clutch performances can influence outcomes in ways that transcend numerical analysis.

## Future Research

Looking ahead, there are plenty of opportunities for future research to build upon our discoveries and tackle the limitations of our study. One promising avenue is to look deeper into the data by analyzing individual player-level statistics. This approach could help us better grasp how specific players contribute to their team's success, providing a more nuanced understanding of the role of each player in three-point shooting dynamics.

Moreover, conducting longitudinal studies that span multiple seasons could offer valuable insights into the evolving trends and patterns of three-point shooting in the NBA. By tracking team performance over time, researchers could uncover long-term dynamics and factors that influence the effectiveness of three-point strategies.

@@

By exploring these avenues of inquiry, future research has the potential to deepen our understanding of the intricate relationship between three-point shooting and team success in the NBA, contributing to advancements in basketball analytics and strategy.

>>>>

---
title: "Beyond the Arc: The Impact of  3 point Shooting in the Modern NBA"
subtitle: "Analysing the relationship between a team’s 3 point shooting and their success in the regular season"
author: 
  - Aryaman Suri
thanks: "Code and data are available at: https://github.com/aryamansuri/NBA-3PointAnalysis"
date: today
date-format: long
abstract: "In recent years, the three-point shot has emerged as a key aspect of basketball strategy in the NBA. Analyzing season-wise NBA data, we assess how teams' three-point performance influences their overall wins. Employing Bayesian Generalized Linear Model, we quantify this relationship, considering factors like team three-point averages across regular seasons. The findings underscore the significance of three-point shooting in NBA success, offering insights into team strategies, player growth, and basketball analytics. Ethical considerations and study constraints are also addressed."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The three point line was introduced in the NBA in the 1979-80 season (@cite-threepoint) and has since altered the nature and balance of the game. Instead of the game being centered around the big man or athletic wings, the 3 point shot gave everybody on the team a new powerful weapon on offenses. Today, there are 155 players who averaged at least one made three pointer in the 2022-23 regular season (@cite-threepoint2).

This paper explores the association between three-point shooting measures and win probability using advanced statistical techniques after analyzing season-wise data. The findings highlight the importance of three-point shooting eﬀiciency as a predictor of team performance in modern basketball and how important it is to be a good three point shooting team for success in today’s game.

Understanding this link has ramifications for numerous parties, including coaches, players, analysts, and team administration. This study adds to the ongoing debate in the NBA about strategic decision-making and player development by explaining the significance of three-point shooting in deciding game outcomes (@cite-debate). Essentially, the question “Does three point shooting actually lead to team success?”, is being asked.

This paper begins by exploring the underlying data in @sec-data, which provides insights into the variables and metrics used in the analyses. Following that, the complexities of the modelling approach are explained in @sec-model, including describing the reasoning behind the statistical techniques and justifying the methodology. The results of the analysis are discussed in @sec-results, diving into the empirical findings and their implications for understanding the relationship between three-point shooting eﬀiciency and NBA team success. @sec-discuss contextualizes the findings of this paper within the larger view of basketball analytics,emphasizing significant takeaways. Finally, the findings and proposed paths for future research are summarized in @sec-discuss2.

# Data {#sec-data}

This study, utilizes  data sourced from the NBA’s official database (@cite-NBA) to examine the relationship between various team performance metrics and their corresponding winning percentages throughout the regular season. The analysis aims to uncover insights into the factors contributing to team success in the NBA, specifically 3 point shooting. The NBA’s database serves as a complete repository of statistical information on teams, players, and games, providing valuable insights into the dynamics of professional basketball. The below sections provide  a detailed overview of the data collection process, cleaning procedures, and subsequent analysis.

@@

We've deliberately opted for **3PM** over **3PA** and **3P%** for several reasons. **3PM** serves as a accurate metric, capturing a team's volume of successful three-point shots with reasonable efficiency. Unlike **3PA**, which might bias the model towards teams overly reliant on three-point attempts irrespective of their success rate, **3PM** provides a balanced representation of both volume and efficiency in three-point shooting. Similarly, **3P%** could skew the analysis towards teams with lower volumes of three-point attempts, overlooking the broader spectrum of team performance.

By prioritizing **3PM** alongside Wins, our model aims to uncover insights into the impact of three-point scoring on overall team success while mitigating potential biases associated with alternative metrics. This strategic selection ensures a throrough exploration of the relationship between three-point performance and winning outcomes in NBA games.

# Model {#sec-model}

@@

I hypothesize a positive relationship between the number of 3-point field goals made by a team and their number of wins throughout the regular NBA season. This hypothesis is rooted in the understanding that teams with higher 3-point shooting efficiency typically demonstrate stronger offensive capabilities, which often translate into higher-scoring games and, consequently, more victories. This relationship is particularly pertinent in today's NBA, where teams increasingly prioritize offensive strategies and exploit the scoring potential offered by 3-point shots.

The chosen Bayesian analysis model (@cite-rstanarm) aims to quantitatively assess the impact of 3-point field goals made on team wins and to provide insights into the significance of this performance metric in determining overall team success. By modeling the relationship between 3-point field goals made and wins using a normal distribution for the response variable, we acknowledge the continuous nature of team wins and seek to capture the variability in win counts across different levels of 3-point shooting proficiency. This approach allows for an understanding of how variations in 3-point performance contribute to variations in team outcomes, thereby elucidating the importance of 3-point shooting in the context of NBA team success. Simply put, teams that excel in scoring 3-point shots are more likely to outscore their opponents and win the games.

# Results {#sec-results}

@@

```{r}
#| echo: false
#| label: fig-2
#| fig-cap: "Scatterplot depicting the **Wins** vs the **3 Points Made** per game"
#| warning: false
ggplot(cleaned_data, aes(x = X3PM, y = W)) +
  geom_point(color = "orange") +
  labs(x = "3PM", y = "Wins", title = "Wins vs 3PM") +
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.grid.major = element_blank(),         # Remove major gridlines
    panel.grid.minor = element_blank()          # Remove minor gridlines
  )
```

@@

team_colors <- c(
  "Atlanta Hawks" = "#E03A3E",
  "Boston Celtics" = "#007A33",
  "Brooklyn Nets" = "#000000",
  "Charlotte Hornets" = "#00788C",
  "Chicago Bulls" = "#CE1141",
  "Cleveland Cavaliers" = "#860038",
  "Dallas Mavericks" = "#00538C",
  "Denver Nuggets" = "#0E2240",
  "Detroit Pistons" = "#C8102E",
  "Golden State Warriors" = "#1D428A",
  "Houston Rockets" = "#CE1141",
  "Indiana Pacers" = "#002D62",
  "LA Clippers" = "#C8102E",
  "Los Angeles Lakers" = "#552583",
  "Memphis Grizzlies" = "#5D76A9",
  "Miami Heat" = "#98002E",
  "Milwaukee Bucks" = "#00471B",
  "Minnesota Timberwolves" = "#0C2340",
  "New Orleans Pelicans" = "#0C2340",
  "New York Knicks" = "#006BB6",
  "Oklahoma City Thunder" = "#007AC1",
  "Orlando Magic" = "#0077C0",
  "Philadelphia 76ers" = "#006BB6",
  "Phoenix Suns" = "#1D1160",
  "Portland Trail Blazers" = "#E03A3E",
  "Sacramento Kings" = "#5A2D81",
  "San Antonio Spurs" = "#C4CED4",
  "Toronto Raptors" = "#CE1141",
  "Utah Jazz" = "#002B5C",
  "Washington Wizards" = "#002B5C"
)
ggplot(grouped_data, aes(x = Average_3PM, y = reorder(TEAM, Average_3PM), fill = TEAM)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Average_3PM, 1)), position = position_nudge(x = 0.5), vjust = 0.5, size = 3, color = "black") +
  labs(x = "Average 3-Pointers Made", y = "Team") +
  scale_fill_manual(values = team_colors) +
  theme(legend.position = "none", axis.text.y = element_text(size = 8))
```
@fig-3 shows the average 3 point shots made per game across the 8 seasons that are involved in our study.
```{r}
#| echo: false
#| label: fig-1
#| fig-cap: "Graph depicting the **Average team wins** and **predicted team wins** vs **Average three point shots made** per game"
#| warning: false
# Calculate predicted wins
predicted_wins <- fitted(nba_model2)

@@

{{< pagebreak >}}

# Discussion {#sec-discuss}

In this study, we aimed to investigate the relationship between three-point shooting efficiency and team success in the NBA. Our analysis involved a detailed examination of season-wise NBA statistics, focusing on variables such as three-pointers made (3PM) and team wins (W). We employed advanced statistical techniques, including data cleaning, correlation analysis, and regression modeling, to uncover insights into the impact of three-point shooting on team performance. By analyzing the data and modeling the relationship between 3PM and W, we sought to show the significance of three-point shooting as a predictor of NBA team success.

Our analysis revealed a compelling association between three-point shooting proficiency and team success in the NBA. We found that teams with higher rates of three-pointers made per game tended to achieve more wins over the course of the regular season. This suggests that three-point shooting plays a key role in determining the outcome of NBA games, with proficient long-range shooters contributing significantly to their team's overall success. These findings underscore the strategic importance of three-point shooting in modern basketball and highlight its impact on team performance.
Additionally, our analysis uncovered insights into the dynamics of three-point shooting and team success. We observed variations in the relationship between 3PM and W across different teams and seasons, indicating that the impact of three-point shooting may vary based on contextual factors such as team composition, playing style, and opponent strategies. Furthermore, our findings suggest that while three-point shooting is a significant predictor of team success, it is not the sole determinant, with other factors such as defense, rebounding, and overall offensive efficiency also influencing and determining game outcomes.

@@

Furthermore, while our modeling approach accounted for various factors, including team performance metrics and game outcomes, it may not capture all relevant variables that could impact the relationship between three-point shooting and team success. Success in the NBA cannot simply be boiled down to efficient volume three point shooting since it is a complex game with many other factors contributing to a team's overall success.

@@

In essence, while the three-point shot remains a potent weapon in the modern NBA, its effectiveness is contingent upon an understanding of its role within a team's overall offensive strategy. As the Houston Rockets learned, basketball is not solely a game of statistics—factors such as matchup dynamics, defensive schemes, and clutch performances can influence outcomes in ways that transcend numerical analysis.

## Future Research {#sec-discuss2}

Looking ahead, there are plenty of opportunities for future research to build upon our discoveries and tackle the limitations of our study. One promising avenue is to look deeper into the data by analyzing individual player-level statistics. This approach could help us better grasp how specific players contribute to their team's success, providing a deeper understanding of the role of each player in three-point shooting dynamics.

Moreover, conducting longitudinal studies that span multiple seasons could offer valuable insights into the trends and patterns of three-point shooting in the NBA. By tracking team performance over time, researchers could uncover long-term dynamics and factors that influence the effectiveness of three-point strategies.

@@

By exploring these avenues of inquiry, future research has the potential to deepen our understanding of the relationship between three-point shooting and team success in the NBA, contributing to advancements in basketball analytics and strategy.

{{< pagebreak >}}
